### 2021-12-03 docs/manuscript/manuscript.tex
Write abstract in a single paragraph and cite "Materials and methods"

\subsection*{Background}
A central problem in developmental and synthetic biology is understanding the
mechanisms by which cells in a tissue or a Petri dish process external cues
and transform such information into a coherent response, e.g., a terminal
differentiation state. It was long believed that this type of positional
information could be entirely attributed to a gradient of concentration of a
specific signaling molecule (i.e., a morphogen). However, advances in experimental
methodologies and computer modeling have demonstrated the crucial role of the
dynamics of a cell's gene regulatory network (GRN) in decoding the information
carried by the morphogen, which is eventually translated into a spatial
pattern. This morphogen interpretation mechanism has gained much attention
in systems biology as a tractable system to investigate the emergent properties
of complex genotype-phenotype maps.

\subsection*{Methods}
In this study, we apply a Markov chain Monte Carlo (MCMC)-like algorithm to probe
the design space of three-node GRNs with the ability to generate a band-like
expression pattern (target phenotype) in the middle of an arrangement of 30
cells, which resemble a simple (1-D) morphogenetic field in a developing embryo.
Unlike most modeling studies published so far, here we explore the
space of GRN topologies with nodes having the potential to perceive
the same input signal differently. This allows for a lot more flexibility during the search
space process, and thus enables us to identify a larger set of potentially
interesting and realizable morphogen interpretation mechanisms.

\subsection*{Results}
Out of 2061 GRNs selected using the search space algorithm, we found 714 classes of
network topologies that could correctly interpret the morphogen.
Notably, the main network motif that generated the target phenotype in
response to the input signal was the type 3 Incoherent Feed-Forward Loop (I3-FFL),
which agrees with previous theoretical expectations and experimental
observations. Particularly, compared to a previously reported pattern forming GRN
topologies, we have uncovered a great variety of novel network designs, some of
which might be worth inquiring through synthetic biology methodologies to test
for the ability of network design with minimal regulatory complexity to
interpret a developmental cue robustly.

@@

Next to topological robustness, we investigated the robustness of the sampled
network topologies to parameter changes (see Methods). Fig~\ref{fig:ab-rob}A
illustrates the association between the abundance of a topology class versus its
robustness. We did not find a significant correlation between
these two variables ($r_S = −0.0389724$, $\text{p-value} = 0.298475$).
However, when we grouped those topologies containing
the same subgraph and calculated their mean robustness, we found that between
the most robust subgraphs is the ``Bistable'' motif, the I3-FFL, as well as
the ``Overlapping domains'' motif (which is a variation of the I3-FFL), which
have been previously reported in Cotterell \& Sharpe~\cite{Cotterell2010}
(see \nameref{S2_Fig} and \nameref{S1_Table}).

>>>>

A central problem in developmental and synthetic biology is understanding the
mechanisms by which cells in a tissue or a Petri dish process external cues
and transform such information into a coherent response, e.g., a terminal
differentiation state. It was long believed that this type of positional
information could be entirely attributed to a gradient of concentration of a
specific signaling molecule (i.e., a morphogen). However, advances in experimental
methodologies and computer modeling have demonstrated the crucial role of the
dynamics of a cell's gene regulatory network (GRN) in decoding the information
carried by the morphogen, which is eventually translated into a spatial
pattern. This morphogen interpretation mechanism has gained much attention
in systems biology as a tractable system to investigate the emergent properties
of complex genotype-phenotype maps. In this study, we apply a Markov chain
Monte Carlo (MCMC)-like algorithm to probe
the design space of three-node GRNs with the ability to generate a band-like
expression pattern (target phenotype) in the middle of an arrangement of 30
cells, which resemble a simple (1-D) morphogenetic field in a developing embryo.
Unlike most modeling studies published so far, here we explore the
space of GRN topologies with nodes having the potential to perceive
the same input signal differently. This allows for a lot more flexibility during the search
space process, and thus enables us to identify a larger set of potentially
interesting and realizable morphogen interpretation mechanisms.
Out of 2061 GRNs selected using the search space algorithm, we found 714 classes of
network topologies that could correctly interpret the morphogen.
Notably, the main network motif that generated the target phenotype in
response to the input signal was the type 3 Incoherent Feed-Forward Loop (I3-FFL),
which agrees with previous theoretical expectations and experimental
observations. Particularly, compared to a previously reported pattern forming GRN
topologies, we have uncovered a great variety of novel network designs, some of
which might be worth inquiring through synthetic biology methodologies to test
for the ability of network design with minimal regulatory complexity to
interpret a developmental cue robustly.

@@

Next to topological robustness, we investigated the robustness of the sampled
network topologies to parameter changes (see Materials and methods).
Fig~\ref{fig:ab-rob}A illustrates the association
between the abundance of a topology class versus its
robustness. We did not find a significant correlation between
these two variables ($r_S = −0.0389724$, $\text{p-value} = 0.298475$).
However, when we grouped those topologies containing
the same subgraph and calculated their mean robustness, we found that between
the most robust subgraphs is the ``Bistable'' motif, the I3-FFL, as well as
the ``Overlapping domains'' motif (which is a variation of the I3-FFL), which
have been previously reported in Cotterell \& Sharpe~\cite{Cotterell2010}
(see \nameref{S2_Fig} and \nameref{S1_Table}).
